Note on Linear Biases in MORUS

### 2018-04-09 morus.tex
minor fix

The trail fragments are depicted on \Cref{fig:minimorusfragments}. In all cases except $\alpha^t_i$, the trail fragment approximates a single AND gate by zero, which holds with probability 3/4, and hence the trail fragment has weight 1. In the case of $\alpha^t_i$, two AND gates are involved; however the two gates share an entry in common, so their sum may be written in the form:
\[
x \cdot y \oplus x \cdot z = x \cdot (y \oplus z).
\]
As a result, the trail fragment $\alpha^t_i$ also has a weight of 1. Another way of looking at this phenomenon is that the trail holds for two different approximations of the AND gates: the alternative approximation is depicted by a dashed line on \Cref{fig:minimorusfragments}.

@@

In more detail, the idea is that by using $\alpha^t_i$, we are able to approximate a bit of $S_0$ using only a ciphertext bit. By combining $\alpha^t_i$ with $\beta^{t+1}_{i+b_0}$, we are then able to approximate a bit of $S_1$ (at step $t+1$) using only ciphertext bits from two consecutive rounds. Likewise, $\gamma^t_i$ allows us to ``jump'' from $S_1$ to $S_4$, i.e. by combining $\alpha^t_i$ with $\beta^t_i$ and $\gamma^t_i$ with appropriate choices of parameters $t$ and $i$ for each, we are able to approximate one bit of $S_4$ using only ciphertext bits. Notice however that $\gamma^t_i$ requires approximating $S_1$ in two consecutive rounds; and so the previous combination requires using $\alpha^t_i$ and $\beta^t_i$ \emph{twice} at different steps. In the same way, $\delta^t_i$ allows us to jump from $S_4$ to $S_2$; and $\varepsilon^t_i$ allows jumping from $S_2$ back to $S_0$. In the end, we are able to approximate a bit of $S_0$ using only ciphertext bits via the combination of all trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$.

However, the same bit of $S_0$ can also be approximated directly by using $\alpha^t_i$ at the corresponding step. Thus that bit can be linearly approximated from two different sides: the first approximation uses a combination of all trail fragments, and involves successive approximations of all state registers (except $S_3$) spanning several encryption steps, as explained in the previous paragraph. The second approximation only involves using $\alpha^t_i$ at the final step reached by the previous trail. By XORing up these two approximations, we are left with only ciphertext bits, spanning five consecutive encryption rounds.

Of course, the overall trail resulting from all of the previous combinations is quite complex, especially since $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$ each require two copies of the preceding trail fragment in consecutive rounds: that is, $\varepsilon^t_i$ requires two approximations of $S_2$, which requires using $\delta^t_i$ twice; and $\delta^t_i$ in turn requires using $\gamma^t_i$ twice, which itself requires using $\alpha^t_i$ and $\beta^t_i$ twice. Then $\alpha^t_i$ is used one final time to close the trail. One may naturally wonder if some components of this trail are in conflict. In particular, products of bits from registers $S_2$ and $S_3$ are approximated multiple times, by $\alpha^t_i$, $\beta^t_i$ and $\gamma^t_i$. To address this concern, and ensure that all approximations along the trail are in fact compatible, we now compute the full trail equation explicitely.

@@

The equation corresponding to each of the five trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, $\varepsilon^t_i$ may be written explicitely as $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ as follows. For each equation, we write on the left-hand side of the equality the biased linear combination used in the trail; and on the right-hand side, the remainder of the equation, which must have non-zero bias (in all cases the bias is $2^{-1}$).
\begin{alignat*}{2}
A^t_i:\quad&& C^t_i \oplus S^{t+1}_{0,i+b_0} &= S^t_{1,i} \oplus S^t_{3,i} \oplus S^t_{1,i} \cdot S^t_{2,i} \oplus S^t_{2,i} \cdot S^t_{3,i}\\
B^t_i:\quad&& C^t_i \oplus S^t_{0,i} \oplus S^t_{1,i} &= S^t_{2,i} \cdot S^t_{3,i} \\
C^t_i:\quad&& S^t_{1,i} \oplus S^{t+1}_{1,i+b_1} \oplus S^t_{4,i} &= S^t_{2,i} \cdot S^t_{3,i}\\
D^t_i:\quad&& S^t_{4,i} \oplus S^{t+1}_{4,i+b_4} \oplus S^{t+1}_{2,i} &= S^t_{0,i} \cdot S^t_{1,i}\\
E^t_i:\quad&& S^t_{2,i} \oplus S^{t+1}_{2,i+b_2} \oplus S^{t+1}_{0,i} &= S^t_{3,i} \cdot S^t_{4,i}
\end{alignat*}

@@

In order to build the equation for the full trail, we start with $E^3_0$:
\[
S^3_{2,0} \oplus S^{4}_{2,b_0} \oplus S^{4}_{0,0} = S^3_{3,0} \cdot S^t_{4,0}.
\]
In order to cancel the $S^{4}_{0,0}$ term on the left-hand side, we add in the equation $A^3_{-b_0}$ (where the sum of two equations of the form $a = b$ and $c = d$ is defined to be $a+c = b+d$). This yields:
\begin{align*}
&S^3_{2,0} \oplus S^{4}_{2,7} \oplus C^{3}_{0,-b_0}\\
=\; &S^3_{3,0} \cdot S^t_{4,0} \oplus S^t_{1,-b_0} \oplus S^t_{3,-b_0} \oplus S^3_{1,-b_0} \cdot S^3_{2,-b_0} \oplus S^3_{2,-b_0} \cdot S^t_{3,-b_0}.
\end{align*}
We then need to cancel two terms of the form $S^t_{2,i}$. To do this, we add in the equations $D^t_i$ for appropriate choices of $t$ and $i$. This replaces the two $S^t_{2,i}$ terms by four $S^t_{4,i}$ terms. By using equation $B^t_i$ four times, we can then replace these four $S^t_{4,i}$ terms by eight $S^t_{1,i}$ terms. By applying equation $B^t_i$ eight times, these eight $S^t_{1,i}$ terms can in turn be replaced by eight $S^t_{0,i}$ terms (and some ciphertext terms). Finally, applying $A^t_i$ eight times allows us to replace these eight $S^t_{0,i}$ terms by only ciphertext bits. Ultimately, for \cipher{MORUS640}, this yields the equation:

@@

That the bias is so high may be suprising: indeed, the full trail uses trail fragment $\varepsilon^t_i$ 1 time, $\delta^t_i$ 2 times, $\gamma^t_i$ 4 times, $\beta^t_i$ 8 times, and $\alpha^t_i$ 9 times. Since each trail fragment has a weight of 1, this would suggest that the total weight should be $1+2+4+8+9 = 24$ rather than 16. However, when combining trail fragments $\beta_i$ and $\gamma_i$, notice that the same AND is computed at the same round between registers $S_2$ and $S_3$. In both cases it is approximated by zero. When XORing the corresponding equations, these two ANDs cancel each other, which saves two AND gates. Since $\gamma^t_i$ is used four times in the course of the full trail, this results in saving 8 AND gates overall, which explains why the final bias is $2^{-16}$ rather than $2^{-24}$.

@@

To confirm that our analysis is correct, we ran experiments on an implementation of \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. In both cases, we measured the bias of two halves $\chi_1$ and $\chi_2$ of the full trail, as well as the bias of the full trail itself (denoted by $\chi$). Results are displayed on \Cref{tab:miniapproximations}. 
While our analysis from the previous section predicts a bias of $2^{-16}$, experiments indicate a slightly higher bias of $2^{-15.5}$ for \cipher{MORUS640}. The discrepancy of $2^{-0.5}$ probably arises from the fact that register bits across different rounds are not completely independent.

@@

To understand why that might be the case, assume a nonlinear term $S^0_{2,0} \cdot S^0_{3,0}$ arises from some part of the trail, and another term $S^0_{2,0} \cdot S^0_{3,w}$ arises from a different part of the trail (where $w$ denotes the word size). Then when we XOR the various trail fragments together, in \cipher{MiniMORUS} these two terms are actually equal and will cancel out, since register-wise rotations by multiple of $w$ bits are ignored. However in the real \cipher{MORUS} these terms are of course distinct and do not cancel out.

@@

In the end, for \cipher{MORUS1280}, since the trail fragment $\gamma^t_i$ is used four times, the phenomenon we just described adds a contribution of $4 \cdot 3 = 12$ to the overall weight of the full trail. This results in a total weight of $4 \cdot 16 + 12 = 76$ (recall that the weight of the trail on \cipher{MiniMORUS1280} is 16). Due to the complexity of the trail, we have confirmed that this bias is correct by explicitely computing the full trail equation in \Cref{sec:traileq}, and computing its exact weight like we did for \cipher{MiniMORUS} in \Cref{sec:minibias}. That is, since the equation is quadratic, we may view it as a graph, which we split into connected components; we then compute the bias of each connected component separately by brute force, and then add up the weights of all components per the piling-up lemma. In the end, the full trail equation does yield a weight of 76 for the full trail on \cipher{MORUS1280}.

In the case of \cipher{MORUS640}, collisions between rotation constants further complicate the analysis. Specifically, when using trail fragment $\beta^t_i$, the term $S^t_{2,i} \cdot S^t_{3,i}$ occurs. As explained previously, a partial collision with the term $S^t_{2,i} \cdot S^t_{3,i-w}$ from trail fragment $\gamma^t_i$ results in \Cref{eq:8circle}. However trail fragment $\alpha^t_{i+d}$ is once used with an offest of $d = b_1+b_4-b_0-b_2$ (relative to $\gamma^t_i$), which in the case \cipher{MORUS640} is equal to $31+13-5-7 = 0 \;\text{mod}\; 32$. This creates another term $S^t_{2,i} \cdot S^t_{3,i}$, which ultimately destroys one of the four occurences of \Cref{eq:8circle}. In the end, when computing the full trail equation on \cipher{MORUS640}, we get that the weight of the trail is 73 (cf. \Cref{sec:traileq}).

>>>>

The trail fragments are depicted on \Cref{fig:minimorusfragments}. In all cases except $\alpha^t_i$, the trail fragment approximates a single AND gate by zero, which holds with probability 3/4, and hence the trail fragment has weight 1. In the case of $\alpha^t_i$, two AND gates are involved; however the two gates share an entry in common, and in both cases the other entry also has a linear contribution to the trail, which results in an overall contribution of the form:
\[
x \cdot y \oplus x \cdot z \oplus y \oplus z= (x \oplus 1) \cdot (y \oplus z).
\]
As a result, the trail fragment $\alpha^t_i$ also has a weight of 1. Another way of looking at this phenomenon is that the trail holds for two different approximations of the AND gates: the alternative approximation is depicted by a dashed line on \Cref{fig:minimorusfragments}.

@@

In more detail, the idea is that by using $\alpha^t_i$, we are able to approximate a bit of $S_0$ using only a ciphertext bit. By combining $\alpha^t_i$ with $\beta^{t+1}_{i+b_0}$, we are then able to approximate a bit of $S_1$ (at step $t+1$) using only ciphertext bits from two consecutive steps. Likewise, $\gamma^t_i$ allows us to ``jump'' from $S_1$ to $S_4$, i.e. by combining $\alpha^t_i$ with $\beta^t_i$ and $\gamma^t_i$ with appropriate choices of parameters $t$ and $i$ for each, we are able to approximate one bit of $S_4$ using only ciphertext bits. Notice however that $\gamma^t_i$ requires approximating $S_1$ in two consecutive steps; and so the previous combination requires using $\alpha^t_i$ and $\beta^t_i$ \emph{twice} at different steps. In the same way, $\delta^t_i$ allows us to jump from $S_4$ to $S_2$; and $\varepsilon^t_i$ allows jumping from $S_2$ back to $S_0$. In the end, we are able to approximate a bit of $S_0$ using only ciphertext bits via the combination of all trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$.

However, the same bit of $S_0$ can also be approximated directly by using $\alpha^t_i$ at the corresponding step. Thus that bit can be linearly approximated from two different sides: the first approximation uses a combination of all trail fragments, and involves successive approximations of all state registers (except $S_3$) spanning several encryption steps, as explained in the previous paragraph. The second approximation only involves using $\alpha^t_i$ at the final step reached by the previous trail. By XORing up these two approximations, we are left with only ciphertext bits, spanning five consecutive encryption steps.

Of course, the overall trail resulting from all of the previous combinations is quite complex, especially since $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$ each require two copies of the preceding trail fragment in consecutive steps: that is, $\varepsilon^t_i$ requires two approximations of $S_2$, which requires using $\delta^t_i$ twice; and $\delta^t_i$ in turn requires using $\gamma^t_i$ twice, which itself requires using $\alpha^t_i$ and $\beta^t_i$ twice. Then $\alpha^t_i$ is used one final time to close the trail. One may naturally wonder if some components of this trail are in conflict. In particular, products of bits from registers $S_2$ and $S_3$ are approximated multiple times, by $\alpha^t_i$, $\beta^t_i$ and $\gamma^t_i$. To address this concern, and ensure that all approximations along the trail are in fact compatible, we now compute the full trail equation explicitely.

@@

The equation corresponding to each of the five trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, $\varepsilon^t_i$ may be written explicitely as $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ as follows. For each equation, we write on the left-hand side of the equality the biased linear combination used in the trail; and on the right-hand side, the remainder of the equation, which must have non-zero bias (in all cases the bias is $2^{-1}$).
\begin{alignat*}{2}
A^t_i:\quad&& C^t_i \oplus S^{t+1}_{0,i+b_0} &= S^t_{1,i} \oplus S^t_{3,i} \oplus S^t_{1,i} \cdot S^t_{2,i} \oplus S^t_{2,i} \cdot S^t_{3,i}\\
B^t_i:\quad&& C^t_i \oplus S^t_{0,i} \oplus S^t_{1,i} &= S^t_{2,i} \cdot S^t_{3,i} \\
C^t_i:\quad&& S^t_{1,i} \oplus S^{t+1}_{1,i+b_1} \oplus S^t_{4,i} &= S^t_{2,i} \cdot S^t_{3,i}\\
D^t_i:\quad&& S^t_{4,i} \oplus S^{t+1}_{4,i+b_4} \oplus S^{t+1}_{2,i} &= S^{t+1}_{0,i} \cdot S^{t+1}_{1,i}\\
E^t_i:\quad&& S^t_{2,i} \oplus S^{t+1}_{2,i+b_2} \oplus S^{t+1}_{0,i} &= S^t_{3,i} \cdot S^t_{4,i}
\end{alignat*}

@@

In order to build the equation for the full trail, we start with $E^2_0$:
\[
S^2_{2,0} \oplus S^{3}_{2,b_2} \oplus S^{3}_{0,0} = S^2_{3,0} \cdot S^2_{4,0}.
\]
In order to cancel the $S^{3}_{0,0}$ term on the left-hand side, we add in the equation $A^2_{-b_0}$ (where the sum of two equations of the form $a = b$ and $c = d$ is defined to be $a+c = b+d$). This yields:
\begin{align*}
&S^2_{2,0} \oplus S^3_{2,b_2} \oplus C^2_{-b_0}\\
=\; &S^2_{3,0} \cdot S^2_{4,0} \oplus S^2_{1,-b_0} \oplus S^2_{3,-b_0} \oplus S^2_{1,-b_0} \cdot S^2_{2,-b_0} \oplus S^2_{2,-b_0} \cdot S^2_{3,-b_0}.
\end{align*}
We then need to cancel two terms of the form $S^t_{2,i}$. To do this, we add in the equations $D^t_i$ for appropriate choices of $t$ and $i$. This replaces the two $S^t_{2,i}$ terms by four $S^t_{4,i}$ terms. By using equation $B^t_i$ four times, we can then replace these four $S^t_{4,i}$ terms by eight $S^t_{1,i}$ terms. By applying equation $B^t_i$ eight times, these eight $S^t_{1,i}$ terms can in turn be replaced by eight $S^t_{0,i}$ terms (and some ciphertext terms). Finally, applying $A^t_i$ eight times allows us to replace these eight $S^t_{0,i}$ terms by only ciphertext bits. Ultimately, for \cipher{MORUS640}, this yields the equation:

@@

That the bias is so high may be suprising: indeed, the full trail uses trail fragment $\varepsilon^t_i$ 1 time, $\delta^t_i$ 2 times, $\gamma^t_i$ 4 times, $\beta^t_i$ 8 times, and $\alpha^t_i$ 9 times. Since each trail fragment has a weight of 1, this would suggest that the total weight should be $1+2+4+8+9 = 24$ rather than 16. However, when combining trail fragments $\beta_i$ and $\gamma_i$, notice that the same AND is computed at the same step between registers $S_2$ and $S_3$ (equivalently, notice that the right-hand side of equations $B^t_i$ and $C^t_i$ is equal). In both cases it is approximated by zero. When XORing the corresponding equations, these two ANDs cancel each other, which saves two AND gates. Since $\gamma^t_i$ is used four times in the course of the full trail, this results in saving 8 AND gates overall, which explains why the final bias is $2^{-16}$ rather than $2^{-24}$.

@@

To confirm that our analysis is correct, we ran experiments on an implementation of \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. In both cases, we measured the bias of two halves $\chi_1$ and $\chi_2$ of the full trail, as well as the bias of the full trail itself (denoted by $\chi$). Results are displayed on \Cref{tab:miniapproximations}. 
While our analysis from the previous section predicts a bias of $2^{-16}$, experiments indicate a slightly higher bias of $2^{-15.5}$ for \cipher{MORUS640}. The discrepancy of $2^{-0.5}$ probably arises from the fact that register bits across different steps are not completely independent.

@@

To understand why that might be the case, assume a nonlinear term $S^0_{2,0} \cdot S^0_{3,0}$ arises from some part of the trail, and another term $S^0_{2,0} \cdot S^0_{3,w}$ arises from a different part of the trail (where $w$ denotes the word size). Then when we XOR the various trail fragments together, in \cipher{MiniMORUS} these two terms are actually equal and will cancel out, since register-wise rotations by multiples of $w$ bits are ignored. However in the real \cipher{MORUS} these terms are of course distinct and do not cancel each other.

@@

In the end, for \cipher{MORUS1280}, since the trail fragment $\gamma^t_i$ is used four times, the phenomenon we just described adds a contribution of $4 \cdot 3 = 12$ to the overall weight of the full trail. This results in a total weight of $4 \cdot 16 + 12 = 76$ (recall that the weight of the trail on \cipher{MiniMORUS1280} is 16). We have confirmed this bias by explicitely computing the full trail equation in \Cref{sec:traileq}, and evaluating its exact weight like we did for \cipher{MiniMORUS} in \Cref{sec:minibias}. That is, since the equation is quadratic, we may view it as a graph, which we split into connected components; we then compute the bias of each connected component separately by brute force, and then add up the weights of all components per the piling-up lemma. In the end, the full trail equation given in \Cref{sec:traileq} yields a weight of 76 for the full trail on \cipher{MORUS1280}.

In the case of \cipher{MORUS640}, collisions between rotation constants further complicate the analysis. Specifically, when using trail fragment $\beta^t_i$, the term $S^t_{2,i} \cdot S^t_{3,i}$ occurs. As explained previously, a partial collision with the term $S^t_{2,i} \cdot S^t_{3,i-w}$ from trail fragment $\gamma^t_i$ results in \Cref{eq:8circle}. However trail fragment $\alpha^t_{i+d}$ is once used in the course of the full trail with an offest of $d = b_1+b_4-b_0-b_2$ (relative to $\gamma^t_i$), which in the case \cipher{MORUS640} is equal to $31+13-5-7 = 0 \;\text{mod}\; 32$. This creates another term $S^t_{2,i} \cdot S^t_{3,i}$, which ultimately destroys one of the four occurences of \Cref{eq:8circle}. In the end, when computing the full trail equation on \cipher{MORUS640}, we get that the weight of the trail is 73 (cf. \Cref{sec:traileq}).
